Union and intersection contracts: theory vs practice

### 2021-04-06 paper.tex
Additonal pass, provide further refs for flow typing

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It turns out these contracts are closely
related to the type casts introduced by gradual typing, modulo blame: both
\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
the authors precisely introduce a system integrating gradual typing with contracts
\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
both statically typed terms, dynamically typed terms, and first-class contracts.
In this setting, each type constructor of the static type system - arrows,
records, foralls - must have a contract counterpart that is able to check at
run-time that a value is indeed a member of the corresponding type.

@@

Products and unions are fundamental and ubiquitous building blocks of program
data. A product is a compound value that hold several pieces of potentially
different types packed together in one value. Dually, a union only holds one
data that may alternatives stored in one value. When nested together, unions and
intersections are able to represent arbitrary tree-like structures, allowing to
model a large class of data.

Unions are also useful for configurations, to represent nullable values in JSON,
plain enumerations, or to support several representations for a field (a file
path could be accepted either as one string of "/"-separated strings, or as a
list of strings).

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors,
together with well-defined semantics. In contrast, in a gradually typed language
like Nickel, the question of the representation of tagged unions as untyped
values arises.

Unfortunately, tagged unions do not have a serializable canonical counterpart as
untyped data. One usual representation is a record with a tag and a value which
shape depends on the tag (see Figure \ref{fig:union-encoding}).

@@

Alas, Nickel does not feature dependent types, which would open a whole
different level of complexity. Choosing the first representation as a default is
not fully satisfying.  First, it does not map precisely to a JSON value: in
consequence, a lot of illegal combinations are structurally valid as untyped
JSON values. But this is to be expected of any representation. More importantly,
this does not necessarily match with the configuration schemas one can find in
the wild, that favor simple, untagged union.

@@

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But as a configuration language, Nickel
does not evolve in the abstract. It aims at the best interoperability possible
with JSON. It is designed to generate data for other softwares to consume, and
tries as much as possible to be expressive enough to encode data schemas at the
language level. If these schemas use untagged union, then this is a strong
incentive to support them at the type level, or at least, at the contract level.

\subsection{Union types}

@@

Untagged union, together with the other existing features of the type system,
are sufficient to implement the encoding mentioned in [cite previous section]:

\begin{lstlisting}
type sum = A of Num | B of Str
\end{lstlisting}

\begin{lstlisting}
{ tag : <A>, val : Num} | { tag : <B>, val : Str}
\end{lstlisting}

@@

Flow typing[ ], as implemented for example by TypeScript, is capable of solving
this issue by correctly propagating the type information in each branch. One can
imagine doing the same in Nickel. Thus equipped with untagged unions and a
typechecker smart enough, classic tagged unions can be simulated, but with the
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of union idioms
appearing in the wild. Thus, adding union types (and their natural and
unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.

>>>>

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Various type constructors have been investigated [All refs], with
solutions of varying complexity. Amongst all the common extensions to the
polymorphic lambda calculus, one of the most simple and ubiquitous turns out to
be one of the most involved: \textit{union types}.

@@

Products and unions are fundamental and ubiquitous building blocks of program
data. A product is a compound data that hold several values of potentially
different shape packed together. Dually, a union only holds one value whose
shape is only known to be in a fixed set of alternatives. When nested together,
unions and intersections are able to represent arbitrary tree-like structures,
allowing to model a large class of data.

Unions are also useful for configurations, to represent nullable values in JSON,
plain enumerations, or to support several representations for a field (for
example, a file path could be accepted either as one string of
\lstinline+"/"+-separated strings, or as a list of strings).

In statically typed languages, and more specifically in functional programming
languages, unions are usually implemented as algebraic data types [references],
also called \textbf{tagged unions}. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors
with well-defined semantics. In contrast, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

Unfortunately, tagged unions do not have a serializable canonical counterpart as
untyped data. One usual representation is a record with a tag and a value whose
shape depends on the tag (see Figure \ref{fig:union-encoding}).

@@

Nickel does not feature dependent types, which would open a whole new level of
complexity. Alas, choosing the non dependent representation as a default is not
fully satisfying.  First, it does not map precisely to a JSON value: in
consequence, a lot of illegal combinations are structurally valid as untyped
JSON values. But this is to be expected of any representation. More importantly,
this does not necessarily match with the configuration schemas one can find in
the wild, that favor simple, untagged union.

@@

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But as a configuration language, Nickel
does not evolve in the abstract. It aims at the best interoperability possible
with JSON. It is designed to generate data for other softwares to consume, and
tries as much as possible to be expressive enough to encode data schemas at the
language level thanks to contracts. If these schemas use untagged union, then
this is a strong incentive to support them at the type level, or at least, at
the contract level.

@@

Untagged union, together with the other existing features of the type system,
are sufficient to implement the encoding illustrated in Figure
\ref{fig:union-encoding}. For example, let us represent an algebraic data type
(written in ML syntax) to an imaginary Nickel union type:

\begin{lstlisting}
type sum = Int of int | Str of string
\end{lstlisting}

\begin{lstlisting}
{ tag : <Int>, val : Num} | { tag : <Str>, val : Str}
\end{lstlisting}

@@

Flow typing, implemented in Racket\cite{FlowTypingRacket1, FlowTypingRacket2},
Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
TypeScript\cite{FlowTypingTypeScript}, solves the issue by being capable of
refining appropriately the type information in each branch. One can imagine
doing the same in Nickel. Thus equipped with untagged unions and a typechecker
smart enough, classic tagged unions can be simulated, but with the additional
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of union idioms
appearing in the wild.  Thus, adding union types to Nickel sounds appealing.
